Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders

### 2025-07-22 Paper/Paper.qmd
update paper

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
    data-sharing: null
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "..."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

In recent years, research on neurodevelopmental disorders (NDDs) has undergone widespread reconsiderations. Traditional diagnostic categories defined in the DSM-5 [@dsm5], have been increasingly called into question in favor of a dimensional and transdiagnostic view [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both]. The DSM-5 itself has incorporated dimensional aspects, for example in defining a broad Autistic Spectrum Disorder (ASD) condition instead of a set of different autism-related diagnoses. Several arguments have contributed to this: the large heterogeneity within diagnostic groups, the poor alignment between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the idea that individuals with NDDs might represent, at least in part, the tails of developmental continua, that are quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

@@

@@

Among the 14 identified articles, six also employed methods of cluster analysis (in all but one case using latent class or latent profile analysis). In five of the six cases, cluster analysis was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was performed as a secondary analysis [@frazier2010asd]. In all cases, clustering results were interpreted as supporting taxonic conclusions, always converging with those from taxometric analyses, with the exception of @frazier2007adhd, where clustering results were partially inconclusive yet suggested the presence of subgroups, in contrast to the dimensional structure indicated by taxometric analysis. In all but two studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.

@@

The present review suggests that taxometric analysis has been underused in neurodevelopmental research, and that many articles using it present relevant methodological limitations or incomplete reporting of crucial information. This is unfortunate in light of the unique relevance of taxometric analysis for the current debate on the categorical vs dimensional nature of NDDs[@astle2022trandiagnostic; @michelini2024world; @happe2021both]. The underuse of taxometric analysis in this field was expected: of the 183 articles employing taxometric methods on psychopathology identified by @haslam2020metaanalysis only 15 were placed in the "childhood" category (encompassing all types, even conditions such as sleep problems).

>>>>

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
    data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "..."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

```{r,message=F,warning=F}
library(readxl)
library(kableExtra)
library(stringr)
library(dplyr)
```

In recent years, research on neurodevelopmental disorders (NDDs) has undergone widespread reconsiderations. Traditional diagnostic categories defined in the DSM-5 [@dsm5], have been increasingly called into question in favor of a dimensional and transdiagnostic view [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial]. The DSM-5 itself has incorporated dimensional aspects, for example in defining a broad Autistic Spectrum Disorder (ASD) condition instead of a set of different autism-related diagnoses. Several arguments have contributed to this: the large heterogeneity within diagnostic groups, the poor alignment between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the idea that individuals with NDDs might represent, at least in part, the tails of developmental continua, that are quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

@@

```{r}
df = data.frame(read_excel("../Literature-review/data/data_extraction.xlsx"))
```

An overview summary of the results (simplified to one row per article) is reported in @tbl-summary. A fully detailed reporting of coded data for all taxometric analyses is available in Supplemental online material. The total number of taxometric analyses was `r nrow(df)`, spanning from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` per article. Targeted disorders were ADHD (6 articles), ASD (6 articles), Dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article targeting both ADHD and ASD. Most articles (10 out of 14) focused on whether disorders were taxa, while the remaining 4 articles targeted whether subtypes were taxa.

The most used taxometric method was MAMBAC, used in 13 out of 14 articles and in `r sum(grepl("MAMBAC",df$Taxometric_methods_used))` out of `r nrow(df)` total analyses, followed by MAXEIG (9 articles), L-Mode (7 articles), MAXCOV (2 articles), and MAXSLOPE (1 article). In most articles more than one taxometric method was used, with a median of `r median(str_count(df$Taxometric_methods_used,";")+1)` methods per analysis. In 11 out of 14 articles the interpretation was based on a quantitative index (Fit~d~ or, for more recent articles, CCFI), while in the remaining 3 cases the interpretation was based on visual inspection of taxometric curves. The number of indicators entered in the analyses spanned from `r min(df$Indicators_number,na.rm=T)` to `r max(df$Indicators_number,na.rm=T)`, with a median of `r median(df$Indicators_number,na.rm=T)` (excluding one article where the number was unclear). The median sample size across all analyses was `r median(df$Sample_size)`, or `r median((df |> group_by(ID_article) |> summarize(avg = mean(Sample_size)))$avg)` after averaging within article.

Taxonic conclusions were mixed. All 5 articles targeting ADHD as a disorder concluded in favor of a dimensional structure, as did both articles targeting SLI. On the contrary, all 4 articles targeting ASD as a disorder leaned towards or clearly supported a categorical (taxonic) structure. The remaining 4 articles targeted subtypes of disorders, and all leaned towards taxonic conclusions, either strongly or predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on Dyslexia subtypes], or at least partly [@ingram2008defining, that supported ASD subtypes as taxa depending on what sets of indicators were used]. In total, 8 out of 14 articles reached taxonic conclusions on at least some of their taxometric analyses (specifically, `r sum(df$Taxonic_conclusion)` out of `r nrow(df)` total analyses, or `r round(mean(df$Taxonic_conclusion)*100,1)`%, reached a taxonic conclusion), mostly concerning ASD (`r sum(grepl("ASD",df$Target_disorder)&df$Taxonic_conclusion==1)` out of `r sum(grepl("ASD",df$Target_disorder))` analyses, or `r round(100*sum(grepl("ASD",df$Target_disorder)&df$Taxonic_conclusion==1)/sum(grepl("ASD",df$Target_disorder)),1)`%).

Artificial admixture was evident or strongly suspected, for at least part of the analyses, in 8 out of 14 articles (in total, involving `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses). Articles where artificial admixture was incurred tended to match those that reached taxonic conclusions, with 6 out of 8 cases. This matching was even higher at the level of individual analysis: taxonic conclusion was reached in `r sum(df$Taxonic_conclusion[df$Artificial_admixture==1])`out of `r sum(df$Artificial_admixture==1)` (`r round(100*mean(df$Taxonic_conclusion[df$Artificial_admixture==1]),1)`%) analyses where admixture was evident or suspected, in contrast to `r sum(df$Taxonic_conclusion[df$Artificial_admixture==0])` out of `r sum(df$Artificial_admixture==0)` (`r round(100*mean(df$Taxonic_conclusion[df$Artificial_admixture==0]),1)`%) analyses where it was probably or certainly excluded.

Remaining methodological issues concerned missing information about descriptive statistics or other indices. In particular, nuisance covariances (i.e., correlations across indicators within diagnostic group or taxa) could be coded for only 2 out of 14 articles. Where reported, it was small to negligible in @james2016latent, while above 0.40 for many indicators in @frazier2023asd; both articles reached a taxonic conclusion. Skewness of distributions of indicators scores was reported by only 5 articles, and where reported it was often close to or beyond $\pm 1$ (at least one indicator in about one third of the analyses); this was not specifically associated to reaching a taxonic conclusion. On the contrary, indicators validity was largely reported, being coded for all but 3 articles, and it was almost always large or very large ($\gg1$).

@tbl-summary *about here*

```{r}
#| label: tbl-summary
#| tbl-cap: Overview summary of the reviewed articles."
summary_tab = read_excel("Tables/Review-Summary.xlsx",sheet=2)
kable(summary_tab, format = "latex", booktabs = TRUE) |>
  kable_styling(font_size = 8, latex_options = c("striped", "scale_down")) |>
  column_spec(1:ncol(summary_tab),width="5em") |>
  landscape()
```

@@

Among the 14 identified articles, 6 also employed methods of cluster analysis (in all but one case using latent class or latent profile analysis). In 5 of the 6 cases, cluster analysis was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was performed as a secondary analysis [@frazier2010asd]. All 6 targeted ADHD and/or ASD, and in all cases clustering results were interpreted as supporting taxonic conclusions, always converging with those from taxometric analyses, with the exception of @frazier2007adhd, where clustering results were partially inconclusive yet suggested the presence of subgroups within ADHD, in contrast to the dimensional structure favored by taxometric analysis. In all but 2 studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.

@@

The first finding of our critical review was that taxometric analysis has generally been underused in neurodevelopmental research, with only 14 articles being identified. This is unfortunate in light of the unique relevance of taxometric analysis for the current debate on the categorical vs dimensional nature of NDDs [@astle2022trandiagnostic; @michelini2024world; @happe2021both; @sonuga2020kuhnian]. The underuse of taxometric analysis in this field was expected: in a meta-analysis conducted by @haslam2020metaanalysis, out 183 articles employing taxometric methods on psychopathology, only 15 were placed in the "childhood" category (this encompassed all types, even conditions such as sleep problems). Considering the studies covered in our review, only ASD, ADHD, and (to a minor extent) SLI were targeted as disorders in taxometric analysis. None of learning disorders was targeted as a taxon, with only Dyslexia being covered in one article, but only for investigating its subtypes.

Despite some heterogeneity of methods, taxometric results were seemingly quite neat: dimensional conclusions were predominantly reached for ADHD and SLI, while categorical/taxonic conclusions were predominantly reached for ASD and wherever subtypes of any disorder were considered. This parallels a conclusion already advanced by @haslam2020metaanalysis, and seems to provide a strong answer to the debate by @chown2021bit and @happe2021both . However, a closer look at the methodological limitations of the reviewed studies warrants more caution in the conclusions.

We identified four studies that targeted ASD as a disorder in taxometric analyses, and all of them reached taxonic conclusions. However, artificial admixture was identified in all of them. Artificial admixture consists of recruiting participants following alternate recruitment channels or criteria, and it is a major threat to conclusion validity, as shown in the above data simulation and in the literature \[...\]. Admixture ranged from sampling participants through different methods, only some of which explicitly targeted disorders [@james2016latent], to admixing clinical and non-clinical samples [@frazier2023asd and @deserno2023probing; in the latter case inclusion criteria from a larger database are partly unclear, but the sample is explicitly structured by diagnosis] and composing a sample including affected and unaffected siblings [@frazier2010asd]. According to our interpretation, the only unbiased taxometric evidence that ASD might be a taxon as a disorder might come from a secondary analysis by @frazier2010asd, where only unaffected siblings were examined, and a taxonic conclusion was still reached (due to the recruitment process, cases with a high level of the trait are oversampled here, but there is no explicit admixture).

Concerning subtypes the situation was similar. All four reviewed articles partly or predominantly reached taxonic conclusions, but three of them probably or mostly incurred admixture. The issue was probably less serious here. In @obrien2012dyslexia, targeting Dyslexia subtypes, we identified only partly risk of admixture, as the authors used psychometric cutoffs on alternative tools and criteria for inclusion. This might not necessarily have affected the results, unless different criteria for inclusion correlate with different underlying deficits (phonological vs non-phonological impairments), however the authors themselves comment in their discussion that: "depending on the dyslexia criterion used, samples may vary in composition with regard to their subtypes" [@obrien2012dyslexia, p.33]

Overall, at the article/sample level there were only 2 major mismatches: @marcus2012latent reached a dimensional conclusions despite incurring admixture by separately recruiting a large number of clinical cases; and @munson2008asd, who reached a taxonic conclusion about ASD subtypes despite no evidence of admixture (all children had received a diagnosis of ASD based on a seemingly uniform clinical assessment).
